\begin{thebibliography}{20}

\bibitem[Gage(1994)]{gage1994}
Gage, P. (1994).
\newblock A new algorithm for data compression.
\newblock \emph{The C Users Journal}, 12(2), 23--38.

\bibitem[Sennrich et al.(2016)]{sennrich2016}
Sennrich, R., Haddow, B., and Birch, A. (2016).
\newblock Neural machine translation of rare words with subword units.
\newblock In \emph{Proceedings of the 54th Annual Meeting of the Association
  for Computational Linguistics (ACL 2016)}, pages 1715--1725.

\bibitem[Schuster and Nakamura(2012)]{schuster2012}
Schuster, M. and Nakamura, K. (2012).
\newblock Japanese and Korean voice search.
\newblock In \emph{2012 IEEE International Conference on Acoustics, Speech
  and Signal Processing (ICASSP)}, pages 5149--5152.

\bibitem[Kudo(2018)]{kudo2018}
Kudo, T. (2018).
\newblock Subword regularization: Improving neural network translation models
  with multiple subword candidates.
\newblock In \emph{Proceedings of ACL 2018}, pages 66--75.

\bibitem[Kudo and Richardson(2018)]{kudo-richardson-2018-sentencepiece}
Kudo, T. and Richardson, J. (2018).
\newblock {S}entence{P}iece: A simple and language independent subword
  tokenizer and detokenizer for neural text processing.
\newblock In \emph{Proceedings of EMNLP 2018: System Demonstrations},
  pages 66--71.

\bibitem[Rust et al.(2021)]{rust2021}
Rust, P., Pfeiffer, J., Vulić, I., Ruder, S., and Gurevych, I. (2021).
\newblock How good is your tokenizer? On the monolingual performance of
  multilingual language models.
\newblock In \emph{Proceedings of ACL 2021}, pages 3118--3135.

\bibitem[Gowda and May(2020)]{gowda2020}
Gowda, T. and May, J. (2020).
\newblock Finding the optimal vocabulary size for neural machine translation.
\newblock In \emph{Findings of EMNLP 2020}, pages 3955--3964.

\bibitem[Murikinati et al.(2020)]{murikinati2020}
Murikinati, P., Anastasopoulos, A., and Neubig, G. (2020).
\newblock Zero-shot cross-lingual transfer of neural machine translation with
  multilingual pretrained encoders.
\newblock In \emph{Proceedings of EMNLP 2020}.

\bibitem[Chakravarthi et al.(2020)]{chakravarthi2020}
Chakravarthi, B.~R., Muralidaran, V., Muthu Kumar, R., and McCrae, J.~P.
  (2020).
\newblock Sentiment analysis of code-mixed {I}ndian languages: An overview of
  {D}ravidian-{C}ode{M}ix@{FIRE} 2020.
\newblock In \emph{Working Notes of FIRE 2020}.

\bibitem[Ranathunga et al.(2019)]{ranathunga2019}
Ranathunga, S., de Silva, N., and Jayasuriya, M. (2019).
\newblock Sinling: {S}inhala natural language processing library.
\newblock In \emph{Proceedings of the 1st Workshop on NLP for Low-Resource
  Underdeveloped Languages}, pages 23--27.

\bibitem[Adelani et al.(2021)]{adelani2021}
Adelani, D.~I. et~al. (2021).
\newblock Masakhaner: Named entity recognition for {A}frican languages.
\newblock \emph{Transactions of the Association for Computational Linguistics},
  9, 1116--1131.

\bibitem[Radford et al.(2019)]{radford2019}
Radford, A., Wu, J., Child, R., Luan, D., Amodei, D., and Sutskever, I.
  (2019).
\newblock Language models are unsupervised multitask learners.
\newblock Technical report, OpenAI.

\bibitem[Kleene(1956)]{kleene1956}
Kleene, S.~C. (1956).
\newblock Representation of events in nerve nets and finite automata.
\newblock In C.~E. Shannon and J.~McCarthy (Eds.), \emph{Automata Studies},
  pages 3--41. Princeton University Press.

\end{thebibliography}
